An Empirical Analysis of the Internet Engineering Task Force with Computational Methods

### 2024-10-02 paper.tex
Remove unneeded methodology section
Update 5.3.3

% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
\subsubsection{Methodology}
\label{sec:org-dyn:hierarchy:methodology}

% From our ICWSM 2024 paper, section 4:

\todo{Do we need a separate methodology section?}

% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
\subsubsection{Centralisation and Dominance}
\label{sec:org-dyn:hierarchy:rq1}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2024/Origin_email_proportions.png}
  \caption{
    Proportion of ``origin" (non reply) emails sent by RPs, WGCs and ADs,
    this should be interpreted in conjunction with
    figure~\ref{fig:activity_proportions} showing the proportion of active
    individuals in each class.
  }
  \label{fig:origin_emails}
\end{figure}

@@

% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
\subsubsection{Communication up and down the Hierarchy}
\label{sec:org-dyn:hierarchy:rq3}

@@

Our third hypothesis is that the cooperative nature of the IETF will be
evident in the effect that power distance~\cite{li2021does} has on
communication throughout the organisation. The raw proportion of the
communication between hierarchy levels helps us to understand the
confidence of lower levels to exercise their voice. This proportion, for a
cooperative and voluntary organisation should be higher for lower hierarchy
levels, showing a confidence in their voice.

We categorise edges based on the hierarchy levels they originate from and
are sent to. Ratios are then calculated of ``upward" versus ``downward"
communication between RPs, WGCs and ADs. We find that the lower the level,
the more of a skew to upward communication.  Communication tends to flow
``up'' the hierarchy more than ``down''.

@@

For \textbf{RQ3}, we determine the direction communication tends to flow
through the hierarchy by looking at inbound and outbound communication
flows. The network's edges are categorised based on their source and target
node's hierarchy level. For each of the inter-level communication
combinations the proportion of communication in the last year going in the
``upward" hierarchy direction is calculated. For instance, if there are six
emails from RPs to WGCs and twelve from WGCs to RPs then the proportion
upward is $\frac{6}{6+12}=\frac{1}{3}$.

In Figure~\ref{fig:inter_strata}, these proportions are plotted, and the
year long window is again pushed forward by one month to increase the
resolution of changes. A proportion of higher than $0.5$ shows more
communication flowing up the hierarchy than down. All three levels are
mostly upward in direction, with the lowest two levels (RP$\rightarrow$
WGC) having the largest skew in the proportion. The other two lines
(RP$\rightarrow$AD and WGC$\rightarrow$AD) show similar periodicity, with
the WGC$\rightarrow$AD line dropping as far below $0.5$ as above showing a
large shift in inter-level communication patterns higher up the hierarchy.

The direct measurement of message flow between the hierarchy levels seems
to bolster our interpretation that the cooperative design of the IETF is
working. This analysis shows that RPs are communicating much more
preferentially upwards in the hierarchy, despite the decrease in overall
share of RP activity. The increase within the WGC level does not have any
noticeable effect on their level of communication, if anything there is a
negative relationship. This again suggests both WGCs and ADs perform a kind
of ``condenser of communication" or ``facilitator" role. In other words,
this analysis suggests the IETF has a ``bottom-up" communication style,
where the higher levels encourage the lower levels' voice.  This is the
desired pattern of communication for a volunteer organisation that is
looking to grow and encourage participants to engage and gradually take on
leadership roles.

Overall for RQ3 then, the pattern observed was communication flowing up the
hierarchy and this corresponds to the answer found in RQ2. WGCs and ADs
were more likely to receive communication with individuals than send
communication to individuals whereas both were more likely than RPs to send
out messages to the list in general.

\begin{figure}[t]
  \centering
  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2024/up_hierarchy_communication_ratios.png}
  \caption{
    Proportion of communications that are ``up" the hierarchy for different
    groups. A proportion above 0.5 indicates that the majority of
    communications are ``upward".
  }
  \label{fig:inter_strata}
\end{figure}

>>>>

% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
% \subsubsection{Methodology}
% \label{sec:org-dyn:hierarchy:methodology}
% 
% % From our ICWSM 2024 paper, section 4:
% 
% \todo{Do we need a separate methodology section?}
% 
% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
\subsubsection{Centralisation and Dominance}
\label{sec:org-dyn:hierarchy:rq1}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2024/Origin_email_proportions.png}
  \caption{
    Proportion of ``origin" (non reply) emails sent by regular
    participants, working group chairs, and area directors, this should be
    interpreted in conjunction with Figure~\ref{fig:activity_proportions}
    showing the proportion of active individuals in each class.
  }
  \label{fig:origin_emails}
\end{figure}

@@

% - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
\subsubsection{Communication Up and Down the Hierarchy}
\label{sec:org-dyn:hierarchy:rq3}

@@

Our third hypothesis is that the cooperative nature of the IETF will be
evident in the effect that power distance~\cite{li2021does} has on
communication. The proportion of the communication between hierarchy levels
helps us to understand the confidence of lower levels to exercise their
voice: that is, the willingness of regular participants to challenge the
working group chairs and of the working group chairs to challenge area
directors.  This proportion, in a cooperative and voluntary organisation,
should be higher for those in lower positions in the hierarchy, showing a
confidence in their voice.

% We categorise edges based on the hierarchy levels they originate from and
% are sent to. Ratios are then calculated of ``upward'' versus ``downward''
% communication between regular participants, working group chairs, and area
% directors. We find that the lower the level, the more of a skew to upward
% communication.  Communication tends to flow ``up'' the hierarchy more than
% ``down''.

@@

We determine the direction communication tends to flow through the
hierarchy by looking at inbound and outbound communication flows. The
network's edges are categorised based on their source and target node's
hierarchy level. For each of the inter-level communication combinations the
proportion of communication in the last year going in the ``upward''
hierarchy direction is calculated. For instance, if there are six emails
from regular participants to working group chairs, and twelve from working
group chairs to regular participants, then the proportion upward is
$\frac{6}{6+12}=\frac{1}{3}$.

In Figure~\ref{fig:inter_strata}, these proportions are plotted, and the
year long window is again pushed forward by one month to increase the
resolution of changes. A proportion of higher than $0.5$ shows more
communication flowing up the hierarchy than down. All three levels are
mostly upward in direction, with the lowest two levels (regular participant
$\rightarrow$ working group chair) having the largest skew in the
proportion. The other two lines (regular participant $\rightarrow$ area
director; working group chair $\rightarrow$ area director) show similar
periodicity, with the working group chairs $\rightarrow$AD line dropping as
far below $0.5$ as above showing a large shift in inter-level communication
patterns higher up the hierarchy.

\begin{figure}[t]
  \centering
  \includegraphics[width=\figureWidthOneColumn]{figures-prev/icwsm-2024/up_hierarchy_communication_ratios.png}
  \caption{
    Proportion of communications that are ``up'' the hierarchy for different
    groups. A proportion above 0.5 indicates that the majority of
    communications are ``upward''.
  }
  \label{fig:inter_strata}
\end{figure}

The direct measurement of message flow between the hierarchy levels seems
to bolster our interpretation that the cooperative design of the IETF is
working. This analysis shows that regular participants are communicating
much more preferentially upwards in the hierarchy, despite the decrease in
overall share of their activity. The increase within the working group
chair level does not have any noticeable effect on their level of
communication, if anything there is a negative relationship. This again
suggests both working group chairs and area directors perform a kind of
``condenser of communication'' or ``facilitator'' role. In other words,
this analysis suggests the IETF has a ``bottom-up'' communication style,
where the higher levels encourage the lower levels' voice.  This is the
desired pattern of communication for a volunteer organisation that is
looking to grow and encourage participants to engage and gradually take on
leadership roles.

\pb{Summary:}
The pattern observed was communication flowing up the hierarchy, broadly
matching that observed in \S\ref{sec:org-dyn:hierarchy:rq2}.  Working group
chairs and area directors were more likely to receive communication with
individuals than send communication to individuals whereas both were more
likely than regular participants to send out messages to the list in
general.
